---
title: "Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("readr")
install.packages("dplyr")
install.packages("ggplot2")

library(readr)
library(ggplot2)
```

```{r}
nacols <- function(df) {
  unlist(lapply(df, function(x) any(is.na(x))))
}

nacols_name <- function(df) {
  colnames(df)[nacols(df)]
}
```

## Problem statement

Given a large data set of client financial information, attempt to determine if a given client will default on their payment.

## Dataset

```{r}
training <- read_csv("./data/cs-training.csv")
training <- training[, -1]

testing <- read_csv("./data/cs-test.csv")
testing <- testing[, -1]
```

Two datasets were used for this study one for training and one for testing with `r nrow(training)` and `r nrow(testing)` entries respectively. Each set includes 10 predictors, with the training set also including a response variable.

The training dataset contains `r sum(nacols(training))` predictors with missing values, namely `r nacols_name(training)` with `r sum(is.na(training))` entries in the dataset containing missing values. Missing values are very prevalent and should be considered in the rest of the study.

